Tropopause Evolution in a Rapidly Intensifying Tropical Cyclone: A Static Stability Budget Analysis in an Idealized, Axisymmetric Framework

### 2018-03-21 paper.tex
Address the rest of John's comments except those on the figures.

Idealized simulations of a TC analyzed by \cite{OhnoSatoh2015} suggested that the development of an upper-level warm core near the 13-km level acted to decrease the static stability near the tropopause within the eye (compare their Figs. 9,10).
Although the mechanisms that might drive this static stability evolution have not been examined explicitly, \cite{SternZhang2013} described the development of the TC warm core using a potential temperature ($\theta$) budget analysis.
They found that radial and vertical advection both played important roles in warm core development throughout RI, and subgrid-scale diffusion became particularly important during the later stage of RI.

Other processes that can modify the static stability in the upper troposphere of TCs include radiative heating within and near the top of the cirrus canopy and shear-induced turbulent mixing near the outflow jet.
To our knowledge, the only paper that has examined explicitly the static stability evolution in a modeled TC is \cite{Kepertetal2016}, but their analysis was limited to the boundary layer.
The analysis herein is based upon that of \cite{SternZhang2013}, except using a static stability budget similar to that of \cite{Kepertetal2016}, with a focus on the upper-tropospheric and lower-stratospheric evolution during RI.

@@

\paragraph{0-24 hours}\mbox{}\\
\indent The initial spin-up period was characterized by a steady increase of the maximum wind speed from 11 m s\textsuperscript{-1} to 22 m s\textsuperscript{-1} (Fig.~\ref{fig:vmax+pmin}a, blue line), an intensification rate that closely matched that of TC Patricia (Fig.~\ref{fig:vmax+pmin}a, red stars).
The weakening of the lower-stratospheric $N^2$ maximum during this period is reflected in the total $N^2$ budget change over this time (Fig.~\ref{fig:stab-00-24}a).
The layer just above the cold-point tropopause was characterized by decreasing $N^2$ (purple shading), maximizing at the storm center.
At and immediately below the tropopause, meanwhile, saw increasing $N^2$ during this time period.
Although these tendencies extended out to the 200-km radius, they were particularly pronounced at innermost radii.
A comparison of the contributions of advection (Fig.~\ref{fig:stab-00-24}b), vertical turbulence (Fig.~\ref{fig:stab-00-24}c), and radiation (Fig.~\ref{fig:stab-00-24}d) reveals that advection was the primary driver of the $N^2$ tendency during this period, acting to stabilize near and just below the tropopause and destabilize above.
Although vertical turbulence acted in opposition to advection (i.e. it acted to stabilize regions that advection acted to destabilize), the magnitude of the advective tendencies was larger, particularly at the innermost radii.
The sum of advection and vertical turbulence (Fig.~\ref{fig:stab-00-24}e) almost exactly replicated the static stability tendencies above the tropopause.
Radiative tendencies, meanwhile, (Fig.~\ref{fig:stab-00-24}d) acted to destabilize the layer below about 16 km and stabilize the layer between 16 and 17 km.
The sum of advection, vertical turbulence, and radiation (Fig.~\ref{fig:stab-00-24}f) reproduced the total change in $N^2$ almost exactly.

@@

Outside of the eye, the $N^2$ evolution exhibited alternating layers of positive and negative tendencies.
Near and above 18 km existed an upward-sloping region of decreasing $N^2$ that extended out to the 180-km radius.
In this region, neither vertical turbulence nor radiation exhibited negative $N^2$ tendencies; advection was the only forcing for this destabilization.
Immediately below this layer, just above the cold-point tropopause, was a region of increasing $N^2$ that sloped upward from 17 km near the 30-km radius to just below 18 km outside of the 100-km radius.
Advection and vertical turbulence both contributed to this positive $N^2$ tendency, with advection playing an important role below about 17.5 km and and turbulence playing an important role above.
The sum of advection and turbulence (Fig.~\ref{fig:stab-24-48}e) reveals two discontiguous regions of increasing $N^2$ in the 17-18-km layer rather than one contiguous region.
The addition of radiation to these two terms, however, (Fig.~\ref{fig:stab-24-48}f) provides the link between these two regions, indicating that radiation also plays a role in strengthening the stable layer just above the tropopause.
In the 16-17-km layer, just below the cold-point tropopause, a horizontally-extensive layer of destabilization also was forced by a combination of advection, vertical turbulence, and radiation.
The sum of advection and vertical turbulence accounts for only a portion of the decreasing $N^2$ in this layer, and actually indicates forcing for stabilization near the 50-km radius and outside of the 130-km radius.
Radiative tendencies overcome this forcing for stabilization in both of these regions to produce the radially-extensive region of destabilization observed just below the tropopause.

@@

  \subsection{The role of radiation}
During the initial spin-up period (0-24 hours; Fig.~\ref{fig:qi+radten}a), convection was not deep enough to deposit large quantities of ice near the tropopause and create a persistent cirrus canopy.
Due to the lack of ice particles, the radiative heating tendencies during this period (Fig.~\ref{fig:qi+radten}b) were relatively small and confined to the region above a few particularly strong, although transient, convective towers.
During RI (24-48 hours), the eyewall updraft strengthened and a radially-extensive cirrus canopy developed near the tropopause (Fig.~\ref{fig:qi+radten}c).
The enhanced vertical gradient of ice mixing ratio at the top of the cirrus canopy induced strong diurnal-mean radiative cooling near the tropopause (Fig.~\ref{fig:qi+radten}d).
This cooling exceeded 0.6 K h\textsuperscript{-1} in some places and sloped downward from the lower stratosphere into the upper troposphere, following the top of the cirrus canopy.
A small radiative warming maximum also appeared outside of the 140-km radius below this region of cooling.
These results broadly agree with those of \citeauthor{buetal} (\citeyear{buetal}; see their Fig. 11a), whose CM1 simulations produced a 0.3 K h\textsuperscript{-1} diurnally-averaged radiative cooling at the top of the cirrus canopy and radiative warming within the cloud that maximized near the 200-km radius.
This broad region of radiative cooling acted to destabilize the layer below the cooling maximum and stabilize the layer above, which can be seen in Fig.~\ref{fig:stab-24-48}d.
The small area of net radiative heating outside of the 140-km radius enhanced the destabilization above 16 km in this region and produced a thin layer of stabilization in the 15-16-km layer.

After the TC's RI period completed (48-72 hours), strong radiative cooling remained near the tropopause at inner radii (Fig.~\ref{fig:qi+radten}f), sloping downward with the top of the cirrus canopy to below the tropopause at outer radii.
Cooling rates exceeded 1 K h\textsuperscript{-1} just above the tropopause between the 30- and 70-km radii.
This value is more than three times the maximum cooling rate of 0.3 K h\textsuperscript{-1} observed by \cite{buetal}, a discrepancy that is a consequence of their larger vertical grid spacing compared to that used here, along with a contribution from differing radiation schemes.
%\cite{buetal} employed the NASA-Goddard radiation scheme in their CM1 simulations, whereas RRTMG is used in the present paper.
To compare our results to those of \cite{buetal}, we ran a simulation identical to that described in Section 2, except using the NASA-Goddard radiation scheme and 625-m vertical grid spacing, to match those of \cite{buetal}.
This simulation produced a maximum 24-hour-average radiative cooling rate of 0.3 K h\textsuperscript{-1}, which agrees with that shown in \cite{buetal}.
Another simulation using 625-m vertical grid spacing and RRTMG radiation produced 24-hour-average cooling rates of up to 0.6 K h\textsuperscript{-1}, which is consistent with the WRF simulations of \cite{buetal}.
This suggests that vertical grid spacing smaller than 625 m is necessary to resolve properly the radiative cooling at the top of the cirrus canopy, and that the results can be quite sensitive to the radiation scheme used.

@@

The results herein suggest that radiative heating tendencies played an important role in destabilizing the upper troposphere and stabilizing the lower stratosphere after the cirrus canopy developed.

@@

%An interesting result of the budget analysis was that turbulent mixing acted to increase static stability in some regions of the storm.
%Although vertical turbulence always acts to eliminate vertical gradients of $\theta$, this adjustment toward a neutral state only occurs where the mixing takes place.
Fig.~\ref{fig:schematic} depicts the effect of turbulent mixing on the $\theta$ profile of an initially stably-stratified layer.
At the initial time in this idealized case, $\theta$ increases with height at a constant rate (Fig.~\ref{fig:schematic}, left panel).
The imposition of tubulence (blue hatching) adjusts the $\theta$ profile within the mixed layer toward a constant value equal to the mean value of that layer in the initial state (Fig.~\ref{fig:schematic}, right panel).
%If turbulence occurs in a stably-stratified layer, it will act to decrease $\theta$ at the top of the layer and increase it at the bottom.
Just above and just below the mixed layer, however, the $\theta$ profile remains undisturbed.
Consequently, although turbulent mixing acts to decrease $\partial \theta/\partial z$ in the layer in which it is occurring, it actually increases $\partial \theta/\partial z$ just below and just above the layer.
These vertical gradients of turbulent mixing are quite important, particularly on the flanks of the upper-tropospheric outflow jet.

Two distinct maxima of vertical eddy diffusivity developed in the tropopause layer as the storm intensified (Fig.~\ref{fig:diff}).
Comparison of these turbulent regions to the $N^2$ tendencies in Figs. \ref{fig:stab-24-48}c and \ref{fig:stab-48-72}c reveals that the layers in which vertical eddy diffusivity maximized corresponded to layers of destabilization due to vertical turbulence.
Just outside of these layers, however, vertical turbulence acted to increase $N^2$.
The large vertical gradient of vertical eddy diffusivity near the tropopause played an important role in developing the lower-stratospheric stable layer during RI.
This supports the hypothesized role of turbulence in setting the outflow-layer $\theta$ stratification in \cite{RotunnoEmanuel}.

@@

The simulated $N^2$ evolution shown herein closely matched that observed during the RI of Hurricane Patricia (2015).
Three processes dominated the $N^2$ variability in the upper troposphere and lower stratosphere: advection, radiation, and vertical turbulence.
Radiation and vertical turbulence played particularly important roles in developing the strong $N^2$ maximum just above the cold-point tropopause during RI.

To put the $N^2$ variability observed near the tropopause into context, Fig.~\ref{fig:fulltrop} depicts the model change in $N^2$ over the RI period from 0 to 21 km altitude, along with the vertical eddy diffusivity and the radiative heating rate.
It is clear that the largest changes in $N^2$ occurred in a relatively shallow layer immediately surrounding the tropopause (Fig.~\ref{fig:fulltrop}a).
This shallow layer also contained the largest diurnally-averaged radiative heating tendencies found anywhere in the domain (Fig.~\ref{fig:fulltrop}c).
%The largest diurnally-averaged radiative heating tendencies found anywhere in the domain also were contained within this relatively shallow layer (Fig.~\ref{fig:fulltrop}c).
Values of vertical eddy diffusivity larger than any found outside of the boundary layer also resided in the upper troposphere (Fig.~\ref{fig:fulltrop}b).
%The upper troposphere also contained a deep layer of vertical eddy diffusivity values larger than those found anywhere outside of the boundary layer.
The results herein suggest that this tubulence not only develops as a response to the presence of small static stability and large vertical wind shear, as discussed by \cite{Molinarietal} and \cite{DuranMolinari2016}, but also can actively increase the static stability in highly localized regions just above and below the mixed layers.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\acknowledgments
We are indebted to George Bryan for his continued development and support of Cloud Model 1.
We also thank Jeffrey Kepert, Robert Fovell, and Erika Navarro for helpful conversations related to this work.
This research was supported by NSF Grant \#1636799.

>>>>

An idealized simulations of a TC analyzed by \cite{OhnoSatoh2015} suggested that the development of an upper-level warm core near the 13-km level acted to decrease the static stability near the tropopause within the eye.
During the early stage of development in their simulation (their Fig. 9), static stability above 16 km was large at all radii.
However, after the storm's intensification, the static stability above 16 km within the eye was markedly smaller (their Fig. 10c).
Although the mechanisms that might drive this static stability evolution have not been examined explicitly it might be related to the development of an upper-tropospheric warm care within the eye.
\cite{SternZhang2013} described the development of the TC warm core using a potential temperature ($\theta$) budget analysis.
Although the warm anomaly in their simulation maximized in the mid-levels, they also note that a secondary warming maximum existed in the 12-14-km layer.
They found that radial and vertical advection both played important roles in warm core development throughout RI, and subgrid-scale diffusion became particularly important during the later stage of RI.
The warming of the upper troposophere by these advective and diffusive processes could contribute to a decrease in static stability near the tropoopause within the eye.
Other processes that can modify the static stability in the upper troposphere of TCs include radiative heating within and near the top of the cirrus canopy and shear-induced turbulent mixing near the outflow jet.

To our knowledge, the only paper that has examined explicitly the static stability evolution in a modeled TC is \cite{Kepertetal2016}, but their analysis was limited to the boundary layer.
The analysis herein is based upon that of \cite{SternZhang2013}, except using a static stability budget similar to that of \cite{Kepertetal2016}, with a focus on the upper-tropospheric and lower-stratospheric evolution during RI.

@@

\paragraph{0-24 hours}\mbox{}\\
\indent The initial spin-up period was characterized by a steady increase of the maximum wind speed from 11 m s\textsuperscript{-1} to 22 m s\textsuperscript{-1} (Fig.~\ref{fig:vmax+pmin}a, blue line), an intensification rate that closely matched that of TC Patricia (Fig.~\ref{fig:vmax+pmin}a, red stars).
The weakening of the lower-stratospheric static stability maximum during this period is reflected in the total $N^2$ budget change over this time (Fig.~\ref{fig:stab-00-24}a).
The layer just above the cold-point tropopause was characterized by decreasing $N^2$ (purple shading), maximizing at the storm center.
At and immediately below the tropopause, meanwhile, existed increasing $N^2$ during this time period.
Although these tendencies extended out to the 200-km radius, they were particularly pronounced at innermost radii.
A comparison of the contributions of advection (Fig.~\ref{fig:stab-00-24}b), vertical turbulence (Fig.~\ref{fig:stab-00-24}c), and radiation (Fig.~\ref{fig:stab-00-24}d) reveals that advection was the primary driver of the $N^2$ tendency during this period, acting to stabilize near and just below the tropopause and destabilize above.
Although vertical turbulence acted in opposition to advection (i.e. it acted to stabilize regions that advection acted to destabilize), the magnitude of the advective tendencies was larger, particularly at the innermost radii.
The sum of advection and vertical turbulence (Fig.~\ref{fig:stab-00-24}e) almost exactly replicated the static stability tendencies above the tropopause.
Radiative tendencies, meanwhile, (Fig.~\ref{fig:stab-00-24}d) acted to destabilize the layer below about 16 km and stabilize the layer between 16 and 17 km.
The sum of advection, vertical turbulence, and radiation (Fig.~\ref{fig:stab-00-24}f) reproduced the total change in $N^2$ almost exactly.

@@

Outside of the eye, the $N^2$ evolution exhibited alternating layers of positive and negative tendencies.
Near and above 18 km existed an upward-sloping region of decreasing $N^2$ that extended out to the 180-km radius.
In this region, neither vertical turbulence nor radiation exhibited negative $N^2$ tendencies; advection was the only forcing for this destabilization.
Immediately below this layer, just above the cold-point tropopause, was a region of increasing $N^2$ that sloped upward from 17 km near the 30-km radius to just below 18 km outside of the 100-km radius.
Advection and vertical turbulence both contributed to this positive $N^2$ tendency, with advection playing an important role below about 17.5 km and and turbulence playing an important role above.
The sum of advection and turbulence (Fig.~\ref{fig:stab-24-48}e) reveals two separate regions of increasing $N^2$ in the 17-18-km layer rather than one contiguous region.
The addition of radiation to these two terms, however, (Fig.~\ref{fig:stab-24-48}f) provides the link between these two regions, indicating that radiation also plays a role in strengthening the stable layer just above the tropopause.
In the 16-17-km layer, just below the cold-point tropopause, a horizontally-extensive layer of destabilization also was forced by a combination of advection, vertical turbulence, and radiation.
The sum of advection and vertical turbulence accounts for only a portion of the decreasing $N^2$ in this layer, and actually indicates forcing for stabilization near the 50-km radius and outside of the 130-km radius.
Radiative tendencies overcome this forcing for stabilization in both of these regions to produce the radially-extensive region of destabilization observed just below the tropopause.

@@

  \subsection{The role of radiation}
During the initial spin-up period (0-24 hours; Fig.~\ref{fig:qi+radten}a), convection was not deep enough to deposit large quantities of ice near the tropopause and create a persistent cirrus canopy.
Due to the lack of ice particles, the radiative heating tendencies during this period (Fig.~\ref{fig:qi+radten}b) were relatively small and confined to the region above a few particularly strong, although transient, convective towers.
During RI (24-48 hours), the eyewall updraft strengthened and a radially-extensive cirrus canopy developed near the tropopause (Fig.~\ref{fig:qi+radten}c).
The enhanced vertical gradient of ice mixing ratio at the top of the cirrus canopy induced strong diurnal-mean radiative cooling near the tropopause (Fig.~\ref{fig:qi+radten}d).
This cooling exceeded 0.6 K h\textsuperscript{-1} (14.4 K day\textsuperscript{-1}) in some places and sloped downward from the lower stratosphere into the upper troposphere, following the top of the cirrus canopy.
A small radiative warming maximum also appeared outside of the 140-km radius below this region of cooling.
These results broadly agree with those of \citeauthor{buetal} (\citeyear{buetal}; see their Fig. 11a), whose CM1 simulations produced a 0.3 K h\textsuperscript{-1} diurnally-averaged radiative cooling at the top of the cirrus canopy and radiative warming within the cloud that maximized near the 200-km radius.
This broad region of radiative cooling acted to destabilize the layer below the cooling maximum and stabilize the layer above, which can be seen in Fig.~\ref{fig:stab-24-48}d.
The small area of net radiative heating outside of the 140-km radius enhanced the destabilization above 16 km in this region and produced a thin layer of stabilization in the 15-16-km layer.

After the TC's RI period completed (48-72 hours), strong radiative cooling remained near the tropopause at inner radii (Fig.~\ref{fig:qi+radten}f), sloping downward with the top of the cirrus canopy to below the tropopause at outer radii.
Cooling rates exceeded 1 K h\textsuperscript{-1} (24 K day\textsuperscript{-1}) just above the tropopause between the 30- and 70-km radii.
This value is more than three times the maximum cooling rate of 0.3 K h\textsuperscript{-1} observed by \cite{buetal}, a difference that is a consequence of their larger vertical grid spacing compared to that used here, along with a contribution from differing radiation schemes.
%\cite{buetal} employed the NASA-Goddard radiation scheme in their CM1 simulations, whereas RRTMG is used in the present paper.
To compare our results to those of \cite{buetal}, we ran a simulation identical to that described in Section 2, except using the NASA-Goddard radiation scheme and 625-m vertical grid spacing, to match those of \cite{buetal}.
This simulation produced a maximum 24-hour-average radiative cooling rate of 0.3 K h\textsuperscript{-1}, which agrees with that shown in \cite{buetal}.
Another simulation using 625-m vertical grid spacing and RRTMG radiation produced 24-hour-average cooling rates of up to 0.6 K h\textsuperscript{-1}, which is consistent with the WRF simulations of \cite{buetal}.
This suggests that vertical grid spacing smaller than 625 m is necessary to resolve properly the radiative cooling at the top of the cirrus canopy, and that the results can be quite sensitive to the radiation scheme used.

@@

The results herein suggest that, after the cirrus canopy developed, radiative heating tendencies considerably destabilized the upper troposphere and stabilized the lower stratosphere.

@@

%An interesting result of the budget analysis was that turbulent mixing acted to increase static stability in some regions of the storm.
%Although vertical turbulence always acts to eliminate vertical gradients of $\theta$, this adjustment toward a neutral state only occurs where the mixing takes place.
Fig.~\ref{fig:schematic} depicts the effect of turbulent mixing on the $\theta$ profile of an initially stably-stratified layer.
At the initial time in this schematic, $\theta$ is assumed to increase with height at a constant rate (Fig.~\ref{fig:schematic}, left panel).
The imposition of turbulence (blue hatching) adjusts the $\theta$ profile within the mixed layer toward a constant value equal to the mean value of that layer in the initial state (Fig.~\ref{fig:schematic}, right panel).
%If turbulence occurs in a stably-stratified layer, it will act to decrease $\theta$ at the top of the layer and increase it at the bottom.
Just above and just below the mixed layer, however, the $\theta$ profile remains undisturbed.
Consequently, although turbulent mixing acts to decrease $\partial \theta/\partial z$ in the layer in which it is occurring, it actually increases $\partial \theta/\partial z$ just below and just above the layer.
These vertical gradients of turbulent mixing are quite important, particularly on the flanks of the upper-tropospheric outflow jet.

Two distinct maxima of vertical eddy diffusivity developed in the tropopause layer as the storm intensified (Fig.~\ref{fig:diff}).
Comparison of these turbulent regions to the $N^2$ tendencies in Figs. \ref{fig:stab-24-48}c and \ref{fig:stab-48-72}c reveals that the layers in which vertical eddy diffusivity maximized corresponded to layers of destabilization due to vertical turbulence.
Just outside of these layers, however, vertical turbulence acted to increase $N^2$.
The large vertical gradient of vertical eddy diffusivity near the tropopause played an important role in developing the lower-stratospheric stable layer during RI.
These results support the hypothesized role of turbulence in setting the outflow-layer $\theta$ stratification in \cite{RotunnoEmanuel}.

@@

The simulated $N^2$ evolution shown herein closely matched that observed during the RI of Hurricane Patricia (2015).
Three $N^2$ budget terms dominated in the upper troposphere and lower stratosphere: advection, radiation, and vertical turbulence.
Radiation and vertical turbulence played particularly important roles in developing the strong $N^2$ maximum just above the cold-point tropopause during RI.

To put the $N^2$ variability observed near the tropopause into context, Fig.~\ref{fig:fulltrop} depicts the model change in $N^2$ over the RI period (hours 24-48) from 0 to 21 km altitude, along with the vertical eddy diffusivity and the radiative heating rate.
The largest changes in $N^2$ occurred in a relatively shallow layer immediately surrounding the tropopause (Fig.~\ref{fig:fulltrop}a).
This shallow layer also contained the largest diurnally-averaged radiative heating tendencies found anywhere in the domain (Fig.~\ref{fig:fulltrop}c).
%The largest diurnally-averaged radiative heating tendencies found anywhere in the domain also were contained within this relatively shallow layer (Fig.~\ref{fig:fulltrop}c).
Values of vertical eddy diffusivity larger than any found outside of the boundary layer also resided in the upper troposphere (Fig.~\ref{fig:fulltrop}b).
%The upper troposphere also contained a deep layer of vertical eddy diffusivity values larger than those found anywhere outside of the boundary layer.
The results herein suggest that this turbulence not only develops as a response to the presence of small static stability and large vertical wind shear, as discussed by \cite{Molinarietal} and \cite{DuranMolinari2016}, but also can actively increase the static stability in highly localized regions just above and below the mixed layers.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ACKNOWLEDGMENTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\acknowledgments
We are indebted to George Bryan for his continued development and support of Cloud Model 1.
We also thank Jeffrey Kepert, Robert Fovell, and Erika Navarro for helpful conversations related to this work.
This research was supported by NSF grant AGS-1636799 and Office of Naval Research Grant N000141712110 as a part of the TCI Departmental Research Initiative.
